# Overview of programming concepts {#programming}

This chapter provides an informal overview of "programming".  Basically, I need to define some terms so that we are all on the same page moving forwards.
This book has no interest in the debates over different programming paradigms: object-oriented vs functional vs whatever.  Rather, we take a practical approach
and move on as quickly as possible.

## What is programming?

Programming is writing down a bunch of instructions into a file so that a computer can execute them.  There really isn't much more to it than that.
Our computers exist to execute the logic that we write down.  If we do a good job, our lives are easier.  If we make a mistake, then we get
the wrong answer. (Telling right from wrong answers coming out of programs is a key skill for you to develop during your PhD.)
If we make a really bad mistake, then we may cause harmful side effects like accidentally deleting data or crashing the machine.
Don't be afraid of mistakes--they are an important part of the process.

### How to learn programming.

Simply put, you will learn the most by doing it.  However, another important way to learn to code is to read code.  Specifically, 
you want to read code that emphasizes the idiomatic way to do things in your language of choice.  Learning the "R" way to do a task
or the "Pythonic" way will help you move from writing code that may be ugly but gets the job done to readable and efficient code 
that you can reuse later.  It is helpful to think of coding as writing, and to remember that writing is re-writing.

If you are entirely new to coding, then I recommend checking out the lessons available from [Software Carpentry](https://software-carpentry.org/).

### Programming at different levels: scripts vs programs

In order to be practical, let's break programming up into two types of tasks.  The first is writing short, focused pieces of code to accomplish
a specific task.  For example, a `bash` script to automate a set of steps to go from raw data to a final set of results.  Another example 
would be writing the code to make a figure or table from the output of the previous `bash` script.  These tasks are often implemented as 
one-off pieces of code that are used for a project and then forgotten about.  (How not to forget about these works, by archiving them
appropriately, is the topic of Chapter \ref{git}).  Colloquially, these short programs are referred to as "scripts" and writing them
as "scripting".  This nomenclature is imperfect, however.

The second type of task is to write a general tool for the research community that accomplishes a specific task.  For example,
a program to align short-read sequence data to a reference genome (@Li2009-th,@Li2013-lh) or a program to simulate genetic
data under explicit evolutionary models [@Kelleher2016-cb].  A related task involves writing a high quality *library* of reusable
code that other researchers may use.  One example of such a product is the [bioconductor](https://www.bioconductor.org/) package for the `R`
programming language.

This class is about the former, although many concepts will apply to the latter.  In general, teaching about the standards required for
a general-purpose tool to have impact is beyond the scope of this course.

## What language should I learn?

Unfortunately, the answer to this question often devolves into "language war" polemics.  "Language *X* sucks because of *Y*", etc..
These arguments are unhelpful beyond getting lots of retweets or visits to your blog.  The reality is that you will probably
be best off with some skills from a few languages rather than a mastery of a single one.  Here, we take a practical approach
and focus on two languages that are popular in "data science" circles.  These are Python and R, which we introduce in
Chapters \ref{Python} and \ref{R}.  I give some guidance regarding language choice beyond these two in Section \ref{languagetypes}
below and in Appendix \ref{otherlanguages}.

## Where to get help?

Like many tasks related to research, programming can be frustrating.  Why does *X* not work, etc.?  Many of you are familiar with online sources of
help, such as [stack overflow](http://www.stackoverflow.com).  These sources tend to be moderated Q&A forums where volunteers respond to
questions from the community.  These resources are invaluable, but are often imperfect.  Answers can be old, out-of-date, or even wrong.
Often, out-of-date and wrong go together, as the answer refers to an old version of the language/library/etc. being asked about.
It can be quite hard for newcomers to tell good from bad answers. It is also important to not get distracted by discussion threads
that go down a rabbit hole of pedantry.  A semi-reliable indicator of a good answer is a question that generated
a lot of conversation resulting in an "accepted" answer with hundreds of votes.  Definitely use these sites, but keep your critical
hat on while looking at the answer.  Don't just copy/paste and pray that your problems have been solves. Rather, try to understand
why the solution works.  And, if you do get a solution there, make a note of it in your code so that future you knows where it came from!

My personal opinion is that good books are invaluable resources. They are often written by language experts and thus focus on the idiomatic
way to do things.  In later chapters, I will give citations to good books and, when possible, give you links to free versions of those books
when they do not violate copyright.

## Basic programming concepts

### Compiled versus interpreted languages {#languagetypes}

It is helpful to subdivide programming languages by what happens to the files you write.   For many languages, your code is
*compiled* from a somewhat human-readable syntax into a machine-readable format that is then executed.  For other languages,
your code is processed by an *interpreter* and then executed.

Python and R, the languages emphasizes here, are interpreted languages, which is important.  First, it means that the speed
at which your code runs *may* be slow relative to the same code written in a compiled language.  However, it is often
faster to write code in interpreted languages.  Such languages often have a richer, more readable syntax than compiled languages,
and they often have more features built-in to their standard libraries (see \ref{stdlib} below).  You also don't have to wait
for your program to compile every time you change something.  Compile times can be substantial for some languages like C++.

For most of your work, I will wager that the performance of interpreted languages is more than sufficient most of the time.  In fact,
the internal guts of most interpreted languages are written in high-performance compiled languages like C when performance
matters.  For example, R's linear algebra support is all written in C, which is completely hidden from the user who only
sees a set of R functions.  In Chapters \ref{Python} and \ref{R}, I'll give some hints about what to do if you run into
performance bottlenecks.

One nice feature of interpreted languages is that they can be used two ways.  You can write all of your code in a file, 
and then run it.  Or, you can use the language's interactive shell environment to enter commands and to exploratory analysis
in real time.  These shells allow the languages to be parts of notebooks, which are the topic of Chapter \ref{notebooks}.

### "Standard" libraries {#stdlib}

The standard library of a language are the features that come with a "plain" installation of the language.  In other words, if
you install R or Python, by definition you get a certain set of features.  Features are added to languages via *libraries*,
which may go by synonyms such as packages, modules, or frameworks, depending on the nomenclature traditions of the language.

A big plus of Python and R is that they have rich standard libraries, which is typical of interpreted languages.  For example,
R has built-in support for high-performance linear algebra operations.  The C programming language, which is compiled, has
no such support, and one needs to depend on third-party packages adding those features, such as the
[GNU Scientific Library](https://www.gnu.org/software/gsl/).

The standard features of a standard library differ from language to language.  There are no rules.  R has built-in support
for graphics, which makes a lot of sense.  R is, after all, primarily a language for statistics.  In Python, you need
[matplotlib](http://www.matplotlib.org) to start making plots.  To get the same linear algebra capabilities in Python that you have
in R, you need [numpy](http://www.numpy.org).

### The components of programming languages {#components}

We now turn to the anatomical elements of programming languages.  While these concepts are pretty universal, we have to pick a language
to use in order to show concrete examples.  I'll use R here, mostly because it has a traditional syntax for writing the code whereas
Python is a bit funky at first.


#### Types {#types}

##### Scalars

##### Arrays

##### Associative Containers

##### Strings

##### Regular expressions, or "regexes"

A large chunk of computing tasks involve text processing.  Sometimes we are interested in matching an exact phrase.
Often, though, we want to do something like find all occurrences of a *pattern* in a string.  For example, find
all words in a document beginning with a capital letter.  These more abstract patterns are found using *regular
expressions*, or *regexes* for short.  Both `R` and Python support searching strings with regexes.  In `R`, 
look to the `stringr` library.  For Python, the relevant module is `re`, which is part of the standard library.
Regexes get complicated fast!  You'll need to practice a lot at first.

#### Variables

Programming languages use variables to store type information and associate it with a name:

```{r}
x <- 5
print(x)
```

Idiomatically, R uses this wacko `<-` syntax for variable assignment.  In fact, this arrow used to be a button on older computers, which is 
where R gets it from.  Pretty much every other language on the planet uses `=` for assignment.  I will admit to never using `<-` in my 
own R code, which will make me the subject of abuse on Twitter if word got out!  There are [cases](https://blog.revolutionanalytics.com/2008/12/use-equals-or-arrow-for-assignment.html)
where it matters which assignment method you use, but I've decided not to worry about it.  Twitter flame wars aside,
this seems like a small point, but it is an example of how there are many nearly equivalent ways of doing the same thing in any language.

#### Keywords

A *keyword* is a fundamental part of a language's syntax.  A keyword may not be used as a variable name because it would
confuse parsing the code

For example, the following block of `R` code are invalid because I am attempting to use a keyword as a variable name:

```{r, eval=F}
if = 4
for = 13245
```

##### "Null" and other special values.

We need a way to express the concept of a variable being undefined.  These "null" values can represent that a variable is undefined, or
has a value that resulted from a domain error in a mathematical calculation.  For example, in R, the NULL value is `NA`:

```{r}
a <- NA
```

There is also a representation of infinity and negative infinity:

```{r}
b <- Inf
c <- -Inf
```

Functions exist to check these values.  In R,

```{R}
is.na(a)
is.na(b)
is.infinite(b)
is.infinite(c)
```

##### Objects

The variable `x` above refers to a simple integer.  In R, it has a `numeric` type:

```{r}
print(class(x))
```

It is common for programming languages to support more complex types, variably referred to as `structs` (structures), `classes`, or some other
term meaning a more complex data type than a simple number or character.  We will introduce these in later chapters as needed, as
the details are language-specific.

#### Mathematical operations

We often need to do math on our numeric variables.  The syntax is usually intuitive:

```{r}
y <- 2
# addition
z <- x + y
# subtraction
z <- x - y
# multiplication
z <- x * y
# division
z <- x/y
```



#### Flow control

For our purposes, computers are simply logic execution boxes.  They do what we tell them to, and we can affect how, when, and
even if a chunk of code is executed via various "flow control" mechanisms.

We can execute code if and only if a certain condition is met:

```{r}
if (x == 5)
{
    print("Yes, x is indeed 5!")
} else {
    print("No, sorry!")
}
```

Languages also provide `for` and `while` loops for flow control:

```{r}
for (i in 1:5)
{
    if (i %% 2 == 0.0)
    {
        print(paste(i,"is even"))
    } else {
        print(paste(i,"is odd"))
    }
}
```

The above is equivalent to:

```{r}
i = 1
while (i < 6)
{
    if (i %% 2 == 0.0)
    {
        print(paste(i,"is even"))
    } else {
        print(paste(i,"is odd"))
    }
    i = i + 1
}
```

By the way, you just learned that `%%` is R's "modulo" operator, which means that `a %% b` returns the remainder of `a/b`.

The various methods of flow control can be mixed, matched, and intermingled.

#### Functions

A fundamental method for organizing code is to write *functions* to perform specific tasks:

```{r}
addtwo <- function(x, y)
{
    return (x + y)
}

x = addtwo(3,-11)
print(x)
```

The way to read the above code is, "There is a function called `addtwo` that takes two arguments, `x` and `y`. The function returns a single value."

Functions are useful because they can make the logic of a program easier to read.  The following is 
pseudocode:

```{r, eval=F}
filename = "datafile.txt"
z = setup(filename)
if (!is.na(z))
{
    process_data(z)
} else {
    stop("Oops--there is an error!")
}
```

#### Comments

All languages support the inclusion of lines of text that are not evaluated as code.  These lines are 
called "comments".  Comments are useful to documenting especially tricky logic, or the fact that
you changed something at a certain time, or where you got the idea for the code from:

```{r}
# The implementation of foo
# inspired by (link to, say, stack overflow)
foo <- function(x)
{
    # ...
}
```

In R and Python, comments begin with a hash sign (\#).

In many languages, include Python and R, it is possible to generate documentation for code directly from the comments.  

#### Error handling

Things go wrong when programs run.  Languages provide means of handling those errors.  Above, we saw
the R function `stop`, which is used to signal that an error has occurred, printing the message provided.

Error handling is an important part of "defensive programming", which I discuss more below.


#### Getting help

R and Python have built-in help systems.  For example, if you don't know what `is.na` does in the above block,
the following command will show you the help page:

```{r, eval=F}
help(is.na)
```

## Programming idioms {#programmingidioms}

### split/apply/combine {#splitapplycombine}

See @Wickham2011-lt for discussion.

## General advice

### RTFM

**R**ead **t**he **F**ull **M**anual.  You'll see this acronym out there a lot, and it is good advice, although
it is crude.  Basically, make sure you read the documentation for the functions you are using.
What does a function do?  What guarantees does it make?  What side effects are there?

An example of where this matters is described [here](https://i-programmer.info/news/231-methodology/13188-python-script-invalidates-hundreds-of-papers.html).  A function in the Python standard library resulted in different answers on different operation systems, causing
a large number of scientific papers to report incorrect results.  The reason is a failure to RTFM.  The method used
to find the files on the computer is documented as returning them in "an arbitrary order", which should have been an immediate
red flag.  Thanks to Edwin Solares for pointing this out to me.

### Write functions

In R and Python, you can write your script "top to bottom" as a list of instructions to execute.  Doing so is *totally fine*!
However, it breaks down the longer your script gets.  Almost any top-to-bottom logic can be rewritten as a set of smaller
functions.  It takes more time, but it forces you to think about what you are doing at each step, it makes error handling
easier, and it makes it easier for your colleagues and future you to understand your code six months later.

### Keep functions short

Ideally, a function does three things:

1. Checks that the arguments passed in are okay.  If not, signal some kind of error.
2. Perform a single task given the input arguments.  If the task goes badly, signal some kind of error
3. Optionally, return some data back to where the function was called from.

This advice is related to some trendy jargon called the "one responsibility principle".

Related advice is to name your functions clearly after what they are doing!  For example, this is unreadable:

```{r}
x <- function(y)
{
}

z <- function(a, b, c)
{
}
```

You need quite a few comments to understand what these functions do!

### Expect errors and test for them

It is really helpful to be able to catch errors where they happen and report the error.  When processing big data sets,
you may come across unexpected input like missing data, etc., that your code doesn't handle.  With time, you'll learn
how to test for these things and automatically write functions that return error messages as early as possible.

When working on shared clusters like the UCI HPC, error checking is paramount!  It is really easy for your work to fail "silently"
because the error message was suppressed.

The next chapters will have more on this topic.

### Editing code

A good "programmer's" text editor is essential.  Like choosing a programming language, editor choice is prone to pointless online
debate.  However, there are some objective truths: word processors are not acceptable!  You **must** use an editor that saves files
in plain text, preferably with Unix newline characters (otherwise your life will be hard on the cluster).  The two classics are 
[emacs](https://www.gnu.org/software/emacs/) and [vim](http://www.vim.org).  I used emacs for about 15 years before pain (google "emacs 
pinky") caused me to switch to vim, and specifically to [neovim](http://neovim.io).

Other options that people like include, but are not limited to:

* [atom](https://atom.io/)
* [Microsoft VS code](https://code.visualstudio.com/)
* [nano](https://www.nano-editor.org/)

### Ergonomics

Make sure you have a proper chair.  Get your monitor set up to the right height.  You probably really like the freedom of a laptop,
but don't forget that it is really bad for you in the long run!  The monitor is making you crane your neck, and the keyboard has you
twisting your wrists at a funny angle.  The taller you are, and the broader your shoulders, the worse all of this is.
All of these issues will catch up to you, eventually. 

This paragraph is one of the few product endorsements in this book.  The [Ergodox EZ](http://www.ergodox-ez.com) keyboard is absolutely
worth the money.  It will, however, slow you down for a while until you get used to the layout.  For portable computing, the 
[Planck EZ](https://ergodox-ez.com/pages/planck) from the same company is great to type with, although you would not expect it based
on its small size.  I've been dealing with carpal tunnel and related issues since about 2001, and these keyboards make a huge difference.
