# The R programming language {#R}

This chapter does not cover the task of programming using R in as much detail as we did
for Python in Chapter \ref{Python}.  The difference does not imply a criticism of R.
Rather, it reflects some practical realities of where my time is best spent and R
will take priority over Python in Chapters \ref{graphics} and \ref{sqlite3}.

## Good books

@Wickham2017-kh is an essential book covering the `tidyverse` (see \ref{tidyverse}).

## Strengths and weaknesses

R is a domain-specific language for statistical analysis. A major strength of R is a pretty full-featured
standard library, usually referred to as `base` R (\ref{base}).

Like Python, a huge array of helpful libraries are available.  Also like Python, it isn't always
easy to discover the one that you need.

Unlike Python, the guidelines for creating R packages for distribution are highly detailed and strictly
adhered to, creating a high average quality of packages installed via [CRAN](http://cran.r-project.org).
Unfortunately, violations of these strict rules can be reported in ways that are often viewed
as unwelcoming and non-inclusive. Certain persons on the R mailing lists are infamous.

Regarding the language itself, most of the weaknesses are found in `base`.  The development of the core
language is highly conservative, meaning that some corners of the language that are klunky have been so
for a long time and will likely be so forever.  You either get used to the oddities or decide to be
in a constant state of vexation.  In many cases, there are libraries on [CRAN](http://cran.r-project.org)
to address some of these issues.

## base {#base}

The standard library excels at the manipulation and processing of rectangular data, meaning "spreadsheet-like"
data arranged into columns and rows.  The reason for this strength is that a date frame type is a core language feature.

R is an excellent platform for all things linear modeling.  There are entire libraries and books on the topic.
In fact, searching Amazon will show you that a rich library exists that covers all sorts of common and esoteric
statistical methods and how to do them in R.

R has built-in support for matrix and vector types as well as linear algebraic operations.  These operations,
and many of the low-level linear regression functions, are implemented in C and are very well-tested.

The standard library, or `base` R (\ref{base}) has good support for plotting, which we discuss more in 
Chapter \ref{graphics}.

Many basic concepts of R were introduced in \ref{components}, so we can focus here on other useful features.

R has five fundamental data structures: vectors, lists, matrices, arrays, and data frames.  It is quite
easy to get some or all of them confused with one another.  If you need to get into the details of these
types for your work, then you need to read about them, and there are many online
tutorials to help you out.

For our purposes, the `data.frame` is where most of the action is:

```{r}
a = 1:4
b = c(TRUE, FALSE, TRUE, FALSE)
c = c("lemming", "potato", "badger", "tomato")

tasty = data.frame(name=c, isanimal=b, tastiness=a)
tasty
```

The `data.frame` is our analog of an in-memory spreadsheet.  We can do group and summarise operations via
the `aggregate` function.  The next code chunk gives us the mean tastiness for each value of `isanimal`:

```{r}
mean_by_isanimal = aggregate(x=tasty[c('tastiness')],
                            by=tasty[c("isanimal")],
                            FUN=mean)
mean_by_isanimal
```

The built in `aggregate` works, but you may find yourself quickly pushing its limits.  I will admit my bias
towards doing these sorts of operations with the `dplyr` library in the so-called `tidyverse`.  The main
reason is for improved performance on large data sets.  The second reason is that you get a toolkit 
for processing databases for free (see Chapter \ref{sqlite3}).

## tidyverse {#tidyverse}

The `tidyverse` is a set of R libraries intended to provide a modernized interface to tasks related to 
the analysis and visualization of data.  It is a collection of many libraries with contributions from many
people. @Wickham2017-kh is an excellent book on these libraries.  A [free version](https://r4ds.had.co.nz/)
is available online from the authors, but it is not identical to the published version.  The chapters are 
in a different order, but most of the content seems to be there.

The `tidyverse` provides plotting capabilities via the very popular `ggplot2` library, which is covered in \ref{ggplot2}.
Other key libraries include `stringr` and `readr`, which you should definitely learn about from the book mentioned above!

The `dplyr` package is an invaluable tool for manipulated `data.frame` and `tibble` objects.  (The latter is
an augmented `data.frame`--see the @Wickham2017-kh book.)

### dplyr

The [dplyr](https://dplyr.tidyverse.org/) library provides a high level grammar for data frame manipulation.
Its interface relies heavily on a new "pipe" operator, `%>%`.  This operator is analogous to the pipe operator 
from the Unix command line (`|`). The origin of the R pipe is the [magrittr](https://magrittr.tidyverse.org/)
library.

Let's just dive into `dplyr` and redo the analysis from the previous section:

```{r}
library(dplyr)
mean_tastiness = tasty %>%
    group_by(isanimal) %>%
    summarise(mt = mean(tastiness))
mean_tastiness
```

In my opinion, the `dplyr` version is much more readable.  I find the "piping" of commands (often called *verbs*)
rather elegant.  There is a lot you can do with `dplyr` for various "data wrangling" tasks, and @Wickham2017-kh
covers many of them. Further, `dpylr` is quite *efficient*, as most of its core operations are implements in C++
using `Rcpp` (see \ref{rcpp}).

A common criticism of the pipe-based syntax is that it is harder to find where errors occur.  I find this to
be a non-issue.  When you do run into problems, just "unpipe" things to debug:

```{r}
g = group_by(tasty, isanimal)
s = summarise(g, mt = mean(tastiness))
s
```

It is less work to semi-unpipe things when debugging:

```{r}
g = tasty %>% group_by(isanimal)
s = g %>% summarise(mt = mean(tastiness))
s
```

These latter two blocks of code are a bit less efficient because they create more *temporary objects* in memory.

## Improving performance {#rcpp}

As with Python, most R code will be fast enough most of the time.  When that is not the case,
[Rcpp](http://www.rcpp.org) is a high-quality library that allows you to write R functions and objects
in modern C++. Rcpp is the most widely depended-upon package in the R software ecosystem.

The official way to get help with Rcpp is via the [mailing list](https://lists.r-forge.r-project.org/cgi-bin/mailman/listinfo/rcpp-devel).
Expect a rebuke if you cross-post to Stack Overflow or other such places.
